---
title: "Faecal sludge truck logistics analysis in Kampala, Uganda"
subtitle: "An exploration to understand the structure and trends of an informal faecal collection sector" 
author:
  name: "Jos van der Ent"
  orcid: 0009-0009-3940-3943
date: today
format: 
  html: 
    number-sections: true
    toc: true
    code-fold: true
    code-folding: true
bibliography: references.bib
execute: 
  echo: true #remove false before submitting as the requirements dictate to add the code to the report
  warning: true
license: "CC BY"
#csl: apa.csl # incase to use custom citation style
---

# Glossary {.unnumbered}

| Abbreviation | Comment       |
|:-------------|---------------|
| FS           | Faecal sludge |

TODO list of used terminology (kable?)



# Introduction

The 'fslogisticskampala' data set is shared on [OpenWashdata](www.openwashdata.org) by Lars Shöbitz. The goal is to provide data sources on faecal sludge transporting logistics in Kampala, Uganda. The data is collected from 30th March 2015 until 25th June 2015 and contains the collection location and dislodge location. By reading the shared dataset and the written report the writer felt it would be interesting to see if there is more information and trends present in the dataset which has not be exposed in the narrative. This dataset is chosen for the Data Science for OpenWashData course. And this specific dataset as it contains location data, in form of GPS coordinates, and multiple datasets which can be joined for the analysis. Additionally it also contains time series which is a common aspect in dataset and so useful for getting hands-on experience

# Methods

## Import libraries

First the required libraries need to be imported to be able to read and process the datasets.

```{r libraries, message = FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gt)
library(here)
```

## Import trips data from fslogisticskampala

The original set has 2 tables @lars_schobitz_fslogisticskampala_2024. One containing information of trips and the second table with metadata on the trucks. First we start with opening the trips.csv

Read trips.csv:

```{r read_trips}
#| message = FALSE 
trips <- read_csv(here::here("data/raw/trips.csv")) |>
  head(4)
```

```{r}
#| message = FALSE 
read_csv(here::here("data/trips_dictionary.csv")) |>
  gt()
```

## Import trucks data from fslogisticskampala

Read trucks.csv:

```{r}
#| message = FALSE 
trucks <- read_csv(here::here("data/raw/trucks.csv")) |>
  head(4)
  
```

```{r}
#| message = FALSE 
read_csv(here::here("data/trucks_dictionary.csv")) |>
gt()
```


## Checking data integrity

In order to understand the data first we will check the raw data.

## Organize data show data pipeline

Cleaning steps

### Write data into folder with cleaned data

trips

```{r}
trips |>
  write_csv(here::here("data/processed/trips.csv"))
```

# Results
```{mermaid}
erDiagram
  CUSTOMER ||--o{ ORDER : places
  ORDER ||--|{ ORDER_ITEM : contains
  PRODUCT ||--o{ ORDER_ITEM : includes
  CUSTOMER {
    string id
    string name
    string email
  }
  ORDER {
    string id
    date orderDate
    string status
  }
  PRODUCT {
    string id
    string name
    float price
  }
  ORDER_ITEM {
    int quantity
    float price
  }
```
## Organise data for analysis

In order to analyse the data additional [@kimball_data_2013]

Generate a date dimension table and dimension time table in @df_star-schema

```{mermaid}
#| label: df_star-schema
#| fig: Start schema for final datafram
flowchart LR
  A[Hard edge] --> B(Round edge)
  B --> C{Decision}
  C --> D[Result one]
  C --> E[Result two]
```

```{mermaid}
flowchart LR
  A[Hard edge] --> B(Round edge)
  B --> C{Decision}
  C --> D[Result one]
  C --> E[Result two]
```
test tex and such

```{mermaid}
erDiagram
  CUSTOMER ||--o{ ORDER : places
  ORDER ||--|{ ORDER_ITEM : contains
  PRODUCT ||--o{ ORDER_ITEM : includes
  CUSTOMER {
    string id
    string name
    string email
  }
  ORDER {
    string id
    date orderDate
    string status
  }
  PRODUCT {
    string id
    string name
    float price
  }
  ORDER_ITEM {
    int quantity
    float price
  }
```

## Explore data

summarise tables map

# Conclusions

# References

::: {#refs}
:::

# License

The original dataset on faecal sludge logistics was published under [CC-BY](https://github.com/openwashdata/fslogisticskampala/blob/main/LICENSE.md) The results of this report are under presented here under similar license:

# Ideas/actions

-   \[\] add license type for research
-   \[\] check weather data
-   \[\] day and night time (dim time)
-   \[\] plot on a map
-   \[\] check for NA values etc
-   \[\] [requirements](https://ds4owd-002.github.io/website/content/project/#tbl-required-items)
