---
title: "Faecal sludge truck logistics analysis in Kampala, Uganda"
subtitle: "An exploration to understand the structure and trends of an informal sector" 
author:
  name: "Jos van der Ent"
date: 18-12-2025
format: 
  html: 
    toc: true
    code-fold: true
    code-folding: true
bibliography: references.bib
---

# Project Description

## Introduction

describe which sources add a thesis or goal of analysis

# Exploring data

## Import libraries

```{r}
library(tidyverse)
```

#TODO: libraries likely needed later: dplyr knitr table library (which?) Geo or mapping library tidyr?

## Import raw data

```{r}
trips <- read_csv(here::here("data/raw/trips.csv"))
```

```{r}
head(trips, 4)
```

```{r}
trucks <- read_csv(here::here("data/raw/trucks.csv"))
```

```{r}
head(trucks, 4)
```

source: 
\#todo fix citations
[@lars_schobitz_data_nodate]

https://openwashdata.github.io/fslogisticskampala/ license: https://github.com/openwashdata/fslogisticskampala/blob/main/LICENSE.md

## Display data

## show data pipeline

# Analysis

# conclusions

# ideas

\[] check weather data 
\[] day and night time 
\[] plot on a map
