---
title: "Faecal sludge truck logistics analysis in Kampala, Uganda"
subtitle: "An exploration to understand the structure and trends of an informal faecal collection sector" 
author:
  name: "Jos van der Ent"
date: 18-12-2025
format: 
  html: 
    number-sections: true
    toc: true
    code-fold: true
    code-folding: true
bibliography: references.bib
---

# Glossary {.unnumbered}

TODO list of used terminology (kable?)

# Introduction

describe which sources add a thesis or goal of analysis

see the published document and think more could be gotten out of the dataset
Personal goal was to master R and use a few specific techniques like linking different data sets, using geographical data and work on time series

# Methods 
Exploring data

## Import libraries
```{r libraries, message = FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

## Import raw data
The original set [@lars_schobitz_data_nodate] has 2 tables. One containing information of trips and the second table with metadata on the trucks 
### Trips
```{r}
#| message = FALSE 
trips <- read_csv(here::here("data/raw/trips.csv")) |>
  head(4)
```

hierboven
```{r, echo = TRUE}
head(trips, 4)
```

```{r}
#| message = FALSE 
trucks <- read_csv(here::here("data/raw/trucks.csv"))
```

```{r}
head(trucks, 4)
```

https://openwashdata.github.io/fslogisticskampala/ license: https://github.com/openwashdata/fslogisticskampala/blob/main/LICENSE.md

## show data pipeline

# Results
## Explore data
summarise tables
map

# Conclusions

# References
::: {#refs}
:::

# Ideas/actions
- add license type for research
- check weather data
- day and night time (dim time)
- plot on a map
-  check for NA values etc
- [requirements](https://ds4owd-002.github.io/website/content/project/#tbl-required-items)
